### 2015-05-23 doc/thesis/typedlua.tex
Adding a statistics summary and a brief review about gradual guarantee

Typed Lua assigns the dynamic type \texttt{any} to the unannotated declarations
that it does not infer a more specific type.
More precisely, Typed Lua does not infer more specific types to the input
parameters of function declarations and to the return type of
recursive functions.
The Typed Lua compiler cannot infer them because it performs type checking in
a single step that simulates the program execution.
We could have split type checking into two steps to try solving this
limitation, but it would have implications in the mechanisms
that Typed Lua uses to handle the discrimination of union types
and the refinement of table types.
We will discuss these features in Section \ref{sec:unions} and
Section \ref{sec:tables}, respectivelly.

@@

This example compiles without any warnings because we annotated
the return type of \texttt{factorial}.
Local type inference cannot use the type of the returned
expression when it includes a function call to the function that is
being type checked, as its return type is still unknown to the type checker.
For this reason, we need to annotate the return type of
recursive local functions to inform the type cheker what type
it should use while type checking their body.

@@

One way to overcome these warnings is to predeclare these functions
with an empty body before the actual declaration.
Even though this is a verbose solution and the Typed Lua compiler
has an option to suppress warnings related to the dynamic type
\texttt{any}, it ensures that further assignments to forwarded
local variables will not change their type.
Next we show how we can apply this alteranative solution to the
previous example, making it compile without any warnings:
\begin{verbatim}
    local function even (n:integer):boolean return true end
    local function odd (n:integer):boolean return false end

@@

Typed Lua type checks this example because the table constructor
has an \emph{unique} table type \texttt{\{"x":1, "y":2\}}
that is consistent with the type in the annotation.
However, if the table constructor had an \emph{open} table type,
it would not type check because it does not allow us to convert
a table field from \texttt{integer} to \texttt{integer?}.
We need this kind of behavior to avoid unsafe constructions
while aliasing \emph{unique} table types, like in the follwoing
example:
\begin{verbatim}
    local t1 = { x = 1 }
    local t2:{"x":integer} = t1
    local t3:{"x":integer?} = t1      --> compile-time warning
    t3.x = nil
\end{verbatim}

>>>>

Typed Lua assigns the dynamic type \texttt{any} to the unannotated declarations
that it does not infer a more specific type.
More precisely, Typed Lua does not infer more specific types to the input
parameters of function declarations and to the return type of
recursive functions.
The Typed Lua compiler cannot infer them because it performs type checking in
a single step that simulates the program execution.
We could have split type checking into two steps to try solving this
limitation, but it would have implications in the mechanisms
that Typed Lua uses to handle the discrimination of union types
and the refinement of table types.
We will discuss these features in Section \ref{sec:unions} and
Section \ref{sec:tables}, respectively.

@@

This example compiles without any warnings because we annotated
the return type of \texttt{factorial}.
Local type inference cannot use the type of the returned
expression when it includes a function call to the function that is
being type checked, as its return type is still unknown to the type checker.
For this reason, we need to annotate the return type of
recursive local functions to inform the type checker what type
it should use while type checking their body.

@@

One way to overcome these warnings is to predeclare these functions
with an empty body before the actual declaration.
Even though this is a verbose solution and the Typed Lua compiler
has an option to suppress warnings related to the dynamic type
\texttt{any}, it ensures that further assignments to forwarded
local variables will not change their type.
Next we show how we can apply this alternative solution to the
previous example, making it compile without any warnings:
\begin{verbatim}
    local function even (n:integer):boolean return true end
    local function odd (n:integer):boolean return false end

@@

Typed Lua type checks this example because the table constructor
has an \emph{unique} table type \texttt{\{"x":1, "y":2\}}
that is consistent with the type in the annotation.
However, if the table constructor had an \emph{open} table type,
it would not type check because it does not allow us to convert
a table field from \texttt{integer} to \texttt{integer?}.
We need this kind of behavior to avoid unsafe constructions
while aliasing \emph{unique} table types, like in the following
example:
\begin{verbatim}
    local t1 = { x = 1 }
    local t2:{"x":integer} = t1
    local t3:{"x":integer?} = t1      --> compile-time warning
    t3.x = nil
\end{verbatim}
